---
title: "Watching Hockey Games While Wearing a Jersey: An Analysis"
author: "Phil Henrickson"
date: "`r Sys.Date()`"
output: html_document
---


```{r}

# knitr chunk options
knitr::opts_chunk$set(echo = T,
                      error=F,
                      warning = F,
                      dev="png",
                      fig.align = 'center',
                      out.width = '80%')

# options for displaying
options(knitr.duplicate.label = "allow",
        scipen = 999)

```


```{r echo = T, include = F}

# packages
library(tidyverse)
library(infer)
library(yardstick)
library(ggthemes)
library(colorr)
library(broom.mixed)


```

```{r}

# ggplot theme
theme_phil <- function () {

        ggplot2::theme_minimal() %+replace%
                theme(
                        axis.title = element_text(),
                        # strip.background = element_rect(fill="grey100"),
                        #strip.text = element_text(colour = 'black'),
                        #  strip.text.x = element_text(size = 7),
                        #  strip.text.y = element_text(size = 7),
                        legend.position = "top",
                        # plot.subtitle = element_text(size = 10,
                        #                              hjust = 0),
                        axis.text.x = element_text(size=rel(0.9),
                                                   vjust = 1),
                        axis.text.y = element_text(size=rel(0.9),
                                                   hjust = 1),
                        plot.caption = element_text(size=8,
                                                    hjust = 1,
                                                    vjust = 1,
                                                    margin = unit(c(6,0,0,0), "points")),
                        legend.title = element_blank(),
                        panel.grid.major = element_blank(),
                        panel.spacing = unit(6.5, "mm"),
                        plot.title = element_text(hjust = 0.5, size = 16),
                        plot.subtitle =  element_text(hjust = 0.5),
                        strip.text.x = element_text(size = 12),
                        panel.grid = element_line(colour = "grey90")

                )
}

# colors for stars games
stars_colors = teamcolors %>%
        filter(league == 'nhl') %>%
        filter(name == 'Dallas Stars')

```

# Data

Load in my jersey labels and Stars game data (from the NHL API) for 22-23, combining these into one dataset.

```{r}

# load in jersey labels
jerseys = data.table::fread(here::here("data", "jerseys", "cleaned_2223.csv"))

# load in games
games = readRDS(here::here("data", "games", "team_games_2223.Rdata")) %>%
        rename(team.id = team.team.id,
               team.name = team.team.name,
               team.link = team.team.link,
               opponent.id = opponent.team.id,
               opponent.name = opponent.team.name,
               opponent.link = opponent.team.link,
               outcome_binary = outcome.team.binary,
               outcome_ordinal = outcome.team.ordinal) %>%
        mutate(team_won = case_when(outcome_binary == 'win' ~ 'yes',
                                    TRUE ~ 'no')) %>%
        set_names(., gsub("\\.", "_", names(.))) %>%
        mutate(date = as.Date(date)) %>%
        as_tibble()

# join up
games_jerseys = bind_cols(jerseys,
                          games) %>%
        mutate(date = as.Date(date)) %>%
        mutate(jersey_worn = case_when(jersey == 'no jersey' ~ 'no',
                                       TRUE ~ 'yes'),
               jersey_type = factor(case_when(grepl("home", jersey) ~ 'home',
                                       grepl("away", jersey) ~ 'away',
                                       TRUE ~ 'no jersey'),
                                    levels = c('no jersey',
                                               'home',
                                               'away'))) %>%
        mutate(jersey = factor(jersey,
                               levels = c("no jersey",
                                          "seguin home",
                                          "pavelski away",
                                          "benn home blackout",
                                          "robertson home rr",
                                          "heiskanen away rr")))

```

# Calendar

This is the calendar of games played.

```{r grid.width = '100%', fig.height =10}

# empty dates for the nhl calendar year
dates = 
        tibble(date = seq(as.Date("2022-10-01"),
                          as.Date("2023-05-31"),
                          by = 1)) %>%
        mutate(Year=lubridate::year(date),
               Month = lubridate::month(date, label=TRUE, abbr=FALSE),
               Day = lubridate::wday(date, label=TRUE),
               mday = lubridate::mday(date),
               Month_week = (5 + day(date) + wday(floor_date(date, 'month'))) %/% 7) %>%
        filter(date < as.Date("2023-04-30")) %>%
        # join with games
        left_join(.,
                  games_jerseys) %>%
        mutate(game_played = case_when(team_won %in% c("yes", "no") ~ 'yes',
                                       TRUE ~ 'no'))


# blank calendar
calendar =         
        dates %>%
        ggplot(aes(y=Month_week, x=Day))+
        geom_tile(color="white")+
        geom_text(aes(label = mday),
                  color = 'white', 
                  size = 8) +
        facet_wrap(Year~Month, ncol=3)+
        scale_y_continuous(breaks = scales::pretty_breaks())+
        scale_y_reverse()+
        theme(legend.position="bottom")+
        theme_phil()+
        ylab("")+
        theme(axis.text.y = element_blank())+
        xlab("")

calendar +
        geom_tile(color = 'white',
                  aes(fill = game_played),
                  linewidth = 4)+
        geom_text(aes(label = mday),
                  color = 'white', 
                  size = 8) +
        scale_fill_manual(values = c("grey80", stars_colors$secondary))

```

These are the games on the calendar broken down by outcome. Victory Green = win; black = loss.

```{r grid.width = '100%', fig.height =10}

calendar +
        geom_tile(color = 'white',
                  aes(fill = team_won),
                  linewidth = 4)+
        geom_text(aes(label = mday),
                  color = 'white', 
                  size = 8)+
        scale_fill_manual(values = c("black", stars_colors$secondary),
                          na.value = 'grey80')

```

The Stars won 47 out of 82 games this year, with a whole heck of a lot of losses occuring in OT.

```{r}

games_jerseys %>%
        group_by(outcome_binary) %>%
        count() %>%
        ggplot(aes(x=outcome_binary,
                   fill = outcome_binary,
                   y = n))+
        geom_col()+
        theme_phil() +
        scale_fill_manual(values = c("black", stars_colors$secondary))

        
games_jerseys %>%
        group_by(outcome_binary) %>%
        count()+
        

```

```{r}

```



I've highlighted in yellow the Stars games in which I wwas wearing a jersey. 

```{r grid.width = '100%', fig.height =10}

calendar +
        geom_tile(color = 'white',
                  aes(fill = team_won),
                  linewidth = 4)+
        geom_tile(data = dates %>%
                          filter(jersey_worn == 'yes'),
                  fill = NA, 
                  color = "yellow", 
                  linewidth = 3) +
        geom_text(aes(label = mday),
                  color = 'white', 
                  size = 8) +
        scale_fill_manual(values = c("grey80", stars_colors$secondary))+
        scale_fill_manual(values = c("black", stars_colors$secondary),
                          na.value = 'grey80')
                                             
```

# Analysis

The basic question: did the Stars win more or less games when I wore a jersey vs when I did not?

```{r}

# bar chart
games_jerseys %>%
        group_by(jersey_worn,
                 team_won) %>%
        count() %>%
        group_by(jersey_worn) %>%
        mutate(prop = n / sum(n)) %>%
        mutate(label = case_when(team_won == 'yes' ~ 'wins',
                                 team_won == 'no' ~ 'losses')) %>%
        ggplot(aes(x=jersey_worn,
                   y = n,
                   fill = team_won))+
        geom_col(color = 'white')+
        geom_text(color = 'white',
                  aes(label = paste(n, label)), position = position_stack(vjust = .5)) +
        scale_fill_manual(values = c('black', stars_colors$secondary))+
        theme_phil()+
        theme(legend.title = element_text())+
        guides(fill = 'none')

```

Turns out they actually won slightly more.




This might be easier to intrepret by looking at the win percentage in games I was wearing a jersey vs games in which I was not.

# proportion chart
games_jerseys %>%
        group_by(jersey_worn,
                 team_won) %>%
        count() %>%
        group_by(jersey_worn) %>%
        mutate(perc = n / sum(n) * 100) %>%
        ggplot(aes(x=jersey_worn,
                   y = perc,
                   fill = team_won))+
        geom_col(color = 'white')+
        geom_text(color = 'white',
                  aes(label = paste(round(perc, 1), "%")), position = position_stack(vjust = .5)) +
        scale_fill_manual(values = c('black', stars_colors$secondary))+
        theme_phil()+
        theme(legend.title = element_text())
```

How do we test this?

Find the difference in proportions between games won when I was wearing a jersey vs games won when I was not.

```{r}

# get observed difference in proportions
d_hat <- games_jerseys %>%
  observe(team_won ~ jersey_worn, success = "yes", 
          stat = "diff in props", order = c("yes", "no"))

# calculate this manually

```

Then, simulate the difference in proportions that we would observe **given that there is no difference between the two groups**. This gives us the range of expected differences in proportions we would observe simply by random chance alone, given our sample.

For example, we can pull out what our split may have looked like in a random sample.

```{r}

# simulate a null distribution
set.seed(1999)
sims_null_dist_d = 
        games_jerseys %>% 
        specify(team_won ~ jersey_worn, success = "yes") %>%
        hypothesize(null = 'independence') %>%
        generate(reps = 10000, type = 'permute')

# one simulation
# most_extremes = 
#         sims_null_dist_d %>%
#         calculate(stat = "diff in props", order = c("yes", "no")) %>%
#         slice_max(stat, n = 5) %>%
#         pull(replicate)

# show some simluations
set.seed(1999)
sample =  sample(1:10000, 1)
        
sims_null_dist_d %>%
        filter(replicate %in% sample) %>%
        group_by(jersey_worn,
                 team_won) %>%
        count() %>%
        group_by(jersey_worn) %>%
        mutate(perc = n / sum(n) * 100) %>%
        ggplot(aes(x=jersey_worn,
                   y = perc,
                   fill = team_won))+
        geom_col(color = 'white',
                 alpha = 0.7)+
        geom_text(color = 'white',
                  aes(label = paste(round(perc, 1), "%")), position = position_stack(vjust = .5)) +
        scale_fill_manual(values = c('black', stars_colors$secondary))+
        theme_phil()+
        theme(legend.title = element_text())+
        ggtitle(paste("Simulated Outcome:", sample))

```

How does our data compare to the simulated distribution? I visualize the distribution of the test statistic in simulations then overlay my observed test statistic. How frequently would we observe a test statistic as extreme as the one I found, in either direction, given that there's no relationship between my wearing a jersey and the outcome? About 33% of the time. In essence, there's nothing particularly notable here.

```{r}

# shade p value
null_dist_d %>%
        calculate(stat = 'diff in props', order = c("yes", "no")) %>%
        visualize() +
        shade_p_value(d_hat,
                      direction = 'two-sided')+
        theme_minimal()

# directly calculate p value
null_dist_d %>%
        calculate(stat = 'diff in props', order = c("yes", "no")) %>%
        get_p_value(d_hat, direction = 'two-sided')

```

In other words, we are well within the realm of what we would expect to see. We are inside what we would expect to see 

```{r}

# get ci
ci = get_confidence_interval(null_dist_d %>%
                                     calculate(stat = 'diff in props', order = c("yes", "no")),
                             level = .95)

# shade p value
null_dist_d %>%
        calculate(stat = 'diff in props', order = c("yes", "no")) %>%
        visualize() +
        shade_confidence_interval(ci)+
        shade_p_value(d_hat,
                      direction = 'two-sided')

```

How many more games would the Stars have needed to win while I was wearing a jersey to flip this around?

Another way of looking at this: how many wins would the Stars have had if I just randomly selected the games in which I was wearing a jersey? This the distribution of wins in those simulations; the red line indicates how many they did in fact win.

```{r}

sims_null_dist_d %>% 
        group_by(replicate) %>% 
        filter(jersey_worn == 'yes' & team_won == 'yes') %>%
        group_by(replicate, team_won) %>% 
        count() %>% 
        ggplot(aes(x=n))+
        geom_bar()+
        geom_vline(xintercept = 25,
                   color = 'red',
                   size = 2)+
        theme_phil()+
        xlab("simulated wins")

```

How many more wins would they have needed while I was wearing my jersey to think there was anything interesting going on here? The 95% confidence interval ranges from 18 wins to 27 wins; if they had won 3 more games things could have been interesting.

```{r}

# at the high end
high_ci = sims_null_dist_d %>% 
    group_by(replicate) %>% 
    filter(jersey_worn == 'yes' & team_won == 'yes') %>%
    group_by(replicate, team_won) %>% 
    count() %>% 
        arrange(desc(n)) %>%
        ungroup() %>% 
        filter(row_number() == 250) %>%
        pull(n)

# at the low end
low_ci = sims_null_dist_d %>% 
    group_by(replicate) %>% 
    filter(jersey_worn == 'yes' & team_won == 'yes') %>%
    group_by(replicate, team_won) %>% 
    count() %>% 
        arrange(desc(n)) %>%
        ungroup() %>% 
        filter(row_number() == 9750) %>%
        pull(n)

c(low_ci, high_ci)

```

Just a note: I was wearing my Heiskanen jersey for the game we lost to the Rangers in OT after giving up a goal at the buzzer, and I was wearing my Robertson jersey for the game we lost to Calgary on a breakaway in the final minute.


```{r}

# calculate diff in props using infer
z_hat <- games_jerseys %>% 
        specify(team_won ~ jersey_worn, success = "yes") %>%
        calculate(stat = "z", order = c("yes", "no"))

# simulate a null distribution
null_dist_z = 
        games_jerseys %>% 
        specify(team_won ~ jersey_worn, success = "yes") %>%
        hypothesize(null = 'independence') %>%
        generate(reps = 1000, type = 'permute') %>%
        calculate(stat = "z", order = c("yes", "no"))

# visualize and add p value
null_dist_z %>%
        visualize() +
        shade_p_value(z_hat,
                      direction = 'both')

# get the p value of a two sided test
null_dist_z %>%
        get_p_value(z_hat,
                    direction = 'two-sided')

# get the p value of a two sided test
null_dist_z %>%
        get_p_value(z_hat,
                    direction = 'both')

```

Okay, but what about by jersey type?

```{r}

games_jerseys %>%
        ggplot(aes(y=jersey,
                   fill = outcome_binary))+
        geom_bar()+
        theme_minimal()+
        scale_fill_colorblind()

```


```{r}

team_season_games %>%
        mutate(game_number = row_number()) %>%
        ggplot(aes(x=game_number, 
                   fill = team.home.away,
                   y= team.shotsOnGoal))+
        geom_col()
      #  scale_x_date(breaks = scales::pretty_breaks())

team_season_games %>%
        ggplot(aes(y = team.home.away,
                   x = team.shotsOnGoal))+
        stat_density_ridges(quantile_lines = T,
                            quantile_fun = median)

library(infer)

# calculate the observed statistic
observed_statistic <- team_season_games %>%
        specify(response = team.shotsOnGoal) %>%
        calculate(stat = "mean")

team_season_games %>%
        ggplot(aes(x=team.shotsOnGoal))+
        geom_histogram(binwidth = 1)

null_dist <- team_season_games %>%
        specify(response = team.shotsOnGoal) %>%
        hypothesize(null = "point", mu = 31) %>%
        generate(reps = 1000, type = "bootstrap") %>%
        calculate(stat = "mean")

null_dist %>%
        visualize() + 
        shade_p_value(observed_statistic,
                      direction = "two-sided")

null_dist %>%
        get_p_value(obs_stat = observed_statistic,
                    direction = "two-sided")

t_test(team_season_games, response = team.shotsOnGoal, mu = 31)

observed_statistic <- team_season_games %>%
        specify(response = team.shotsOnGoal) %>%
        hypothesize(null = "point", mu = 31) %>%
        calculate(stat = "t") %>%
        dplyr::pull()

null_dist <- team_season_games %>%
        specify(team.shotsOnGoal ~ team.home.away) %>%
        generate(reps = 1000, type = "bootstrap") %>%
        calculate(stat = "t", order = c("home", "away"))

observed_stat <- team_season_games %>%
        specify(team.shotsOnGoal ~ team.home.away) %>%
        calculate(stat = "t", order = c("home", "away"))

null_dist %>%
        visualise(stat = 't')+
        shade_p_value(observed_statistic,
                      direction = "two-sided")

lm(team.shotsOnGoal ~ team.home.away, data = team_season_games) %>%
        broom::tidy() %>% 
        mutate_if(is.numeric, round, 2)


# 
# team_season_games %>%
#         specify(team.shotsOnGoal ~ team.home.away) %>%
#         hypothesize(null = "independence") %>%
#         generate(reps = 1000, type = "bootstrap") %>%
#         calculate(stat = "diff in means", order = c("home", "away")) %>%
#         visualise()+
#         shade_p_value(observed_statistic,
#                       direction = "two-sided")


```




```{r}


games_jerseys %>%
        group_by(jersey_worn, team_won) %>%
        count()

games_jerseys %>%
        ggplot(aes(x=jersey_worn,
                   fill = team_won))+
        geom_bar(position = 'fill')+
        theme_minimal()

games_jerseys %>%
        ggplot(aes(y=jersey_worn,
                   fill = outcome_binary))+
        geom_bar()+
        theme_minimal()+
        scale_fill_colorblind()
        theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

games_jerseys %>%
        ggplot(aes(y=jersey,
                   fill = outcome_binary))+
        geom_bar()+
        theme_minimal()+
        scale_fill_colorblind()

```


test

```{r}


d_hat <- games_jerseys %>% 
        specify(outcome_binary ~ jersey_worn, success = "win") %>%
        calculate(stat = "diff in props", order = c("yes", "no"))

null_dist <- games_jerseys %>% 
        specify(outcome_binary ~ jersey_worn, success = "win") %>%
        hypothesize(null = 'independence') %>%
        generate(reps = 1000, type = 'permute') %>%
        calculate(stat = "diff in props", order = c("yes", "no"))

null_dist %>%
        visualise() +
        shade_p_value(d_hat, direction = 'two sided')

null_dist %>%
  get_p_value(obs_stat = d_hat, direction = "two sided")

```
```{r}

games_jerseys %>% 
        specify(outcome_binary ~ jersey_worn, success = "win") %>%
        fit()

```



```{r}

glm(team_won ~ jersey, 
    data = games_jerseys %>%
            mutate(team_won = case_when(team_won == 'yes' ~ 1,
                                        TRUE ~ 0)),
    family = "binomial") %>%
        summary()

glm(team_won ~ jersey_worn + jersey_type, 
    data = games_jerseys %>%
            mutate(team_ = case_when(team_won == 'yes' ~ 1,
                                        TRUE ~ 0)),
        family = "binomial") %>%
        tidy() %>%
        mutate_if(is.numeric, round, 3)



fit = lm(team_won ~ jersey, 
    data = games_jerseys %>%
            mutate(team_won = case_when(team_won == 'yes' ~ 1,
                                        TRUE ~ 0))) 

# white robust standard errors
library(lmtest)
library(sandwich)
coeftest(fit, vcov = vcovHC(fit, type = 'HC0')) %>%
        tidy() %>%
        mutate_if(is.numeric, round, 3)


# get bootstrapped standard errors
rsample::bootstraps(games_jerseys %>%
                  select(team_won, outcome_ordinal, outcome_binary, jersey_type, jersey_worn) %>%
                  as_tibble())


bootstrap(games_jerseys %>%
                  select(team_won, outcome_ordinal, outcome_binary, jersey_type, jersey_worn) %>%
                  as_tibble(),
          m = 100)
games_jerseys %>%
            mutate(team_won = case_when(team_won == 'yes' ~ 1,
                                        TRUE ~ 0)))

```


